Time and space specifications for stencils

### 2016-06-16 paper.tex
amplifying informal intuition on inference procedure

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}#1\\#2\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

@@

\paragraph{Basic syntax definitions} In the following, $v$
  ranges over variables, $s$ over statements, and $e$ over expressions
  (which may be impure).  By ``variable'', we mean in the imperative
  sense (\ie{}, named binders to memory cells) instead of mathematical
  variables. The function \rhsExp{} maps statements to a set of
  expressions which occur in right-hand positions (\ie{}, not the
  target of an assignment). The function \var{} maps expressions to a
  set of the variables used in right-hand positions.

@@

Let $M$ range over the finite maps generated by \textsf{analyse}.  For
each $v \in \mathsf{dom}(M)$, the next step of the algorithm coalesces
the set of relative index vectors $M(v)$ into contiguous regions within an
$n$-dimensional space.

@@

\begin{definition}[Span]
  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vecvtors. 
\end{definition}

Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \; 
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional spaace. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

@@

\item Sort each permutation set into an ordered list, based on the ordering:
\[
\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i 
\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
\]
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0} 
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0} 
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial $\bullet$ which 
coalescees and contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}, 
\stwo{0}{1}{0}{1}, 
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
 \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) = 
\{\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}
\stwo{0}{1}{0}{1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{-1}{0}{1}
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Filter by $\containedin$, defined: 
%
\begin{align*}
(\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = 
\vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
\end{align*}

Yields:
\begin{align*}
\stwo{-1}{0}{1}{0}, 
\stwo{0}{-1}{0}{1}
\end{align*}
\end{enumerate}

\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (2,2);
\node at (1.25,1.22) {i, j};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
%
\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (2,2);
\node at (1.25,1.22) {i, j};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
%
\begin{tikzpicture}
\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\draw[->] (0.5,2) -- (0.5,0.5);
\node at (1.25,1.22) {i, j};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}

\subsection{Infering temporal specifications}

\begin{itemize}
\item Extract from syntax finite map of variables
to relative-index verctors representing the contributing
array subscritps
\end{itemize}

>>>>

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

@@

\renewcommand*{\arraystretch}{0.8}
\paragraph{Basic syntax definitions} In the following, $v$
  ranges over variables, $s$ over statements, and $e$ over expressions
  (which may be impure).  By ``variable'', we mean in the imperative
  sense (\ie{}, named binders to memory cells) instead of mathematical
  variables. The function \rhsExp{} maps statements to a set of
  expressions which occur in right-hand positions (\ie{}, not the
  target of an assignment). The function \var{} maps expressions to a
  set of the variables used in right-hand positions.

@@

Let $M$ range over the finite maps generated by \textsf{analyse}.  For
each $v \in \mathsf{dom}(M)$, the algorithm then constructs a group
of regions which cover all contiguous groups of relative indices
(from $M(v)$) in the $n$-dimensional space.

Informally, the procedure proceeds as follows. First, relative
indices are turned into unit regions in $n$-dimensions. For
example, for the five point stencil there are five $1 \times 1$ regions,
called \emph{spans}, which are essentially $0$-dimensional:
%
\begin{center}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,1) rectangle (1.5,1.5);
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\end{center}
%
Each dimension is traversed, coalescing consecutive spans
which vary in $m$-dimensions into coalesced spans which vary in 
$m+1$-dimensions. In this example, the $0$-dimensional spans
 become $1$-dimensional spans (rows/columns):
%
\begin{align*}
\begin{array}{cc}
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (0.5,1) rectangle (2,1.5);
\draw[thick] (1,0.5) rectangle (1.5,1);
\draw[thick] (1,1.5) rectangle (1.5,2);
\end{tikzpicture}
\qquad
&
\qquad
\begin{tikzpicture}
%\draw[step=0.5cm,lightgray,very thin] (0.5,0.5) grid (2,2);
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1.5,1) rectangle (2,1.5);
\draw[thick] (0.5,1) rectangle (1,1.5);
\draw[thick] (1,0.5) rectangle (1.5,2);
\end{tikzpicture}
\\
\textit{dimension 1} \qquad & \qquad \textit{dimension 2}
\end{array}
\end{align*}
%
Any spans that are contained with any other span are deleted,
leaving a minimal set of spans (which may overlap, but none of which
fully contains another).
\begin{center}
\begin{tikzpicture}
%\node at (1.25,1.22) {\textnormal{\small{0,0}}};
\draw[thick] (1,0.5) rectangle (1.5,2);
\node at (2,1.22) {+};
\draw[thick] (2.5,1) rectangle (4,1.5);
%\node at (2.75,1.22) {\textnormal{\small{0,0}}};
\end{tikzpicture}
\end{center}
%
The procedure is then iterated till a fixed point is reached. In this
example this is reached in the first step. These final spans
go on to form the basis of the spatial stencil specification, where
each span is a region product \texttt{*} and multiple spans are
combined with the region sum \texttt{+}.

\paragraph{Span coalescing, formally}

@@

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vecvtors. 
\end{definition}

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows. 
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \; 
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional spaace. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

@@

\item Sort each permutation set into an ordered list, based on the ordering:
\[
\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i 
\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
\]
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0} 
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0} 
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial $\bullet$ which 
coalescees and contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}, 
\stwo{0}{1}{0}{1}, 
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
 \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) = 
\{\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}
\stwo{0}{1}{0}{1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{-1}{0}{1}
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
\end{align*}
For our example this then yields the final result of: 
\begin{align*}
\textsf{regions}(M) 
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$. 
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.
